Quantitative neuroimaging with ANTs

### 2017-04-24 ibm.tex
ENH:  ASNR 2017

\documentclass[ignorenonframetext,]{beamer}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
\usetheme{AnnArbor}
\usecolortheme{dolphin}
\usefonttheme{professionalfonts}
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\newif\ifbibliography
\usepackage{longtable,booktabs}
\usepackage{caption}
% These lines are needed to make table captions work with longtable:
\makeatletter
\def\fnum@table{\tablename~\thetable}
\makeatother
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight0.8\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}

@@

\section{Developers and
collaborators}\label{developers-and-collaborators}

\includegraphics{./figures/brian_and_nick.png}

\begin{frame}

\includegraphics{./figures/antsCollaborators2.png}

@@

\section{Why would you care?}\label{why-would-you-care}

\begin{frame}{Software for medical image analysis}

\begin{itemize}
\item
  \href{http://fsl.fmrib.ox.ac.uk/fsldownloads/}{FSL}
\item
  \href{http://www.fil.ion.ucl.ac.uk/spm/software/}{SPM}
\item
  \href{http://www.fil.ion.ucl.ac.uk/spm/software/}{FreeSurfer}
\item
  \href{http://mipav.cit.nih.gov}{MIPAV}
\item
  \href{https://afni.nimh.nih.gov/afni}{AFNI}
\item
  Slicer, Elastix, SimpleITK, ANTs \(\longleftrightarrow\)
  \href{http://www.itk.org}{Insight Toolkit}
\item
  Many more at \href{http://idoimaging.com}{idoimaging.com}
\end{itemize}

@@

\begin{frame}{International competitions}

\begin{itemize}
\item
  \href{http://www.ncbi.nlm.nih.gov/pubmed/19195496}{Klein 2009}: MRI
  brain registration
\item
  \href{http://empire10.isi.uu.nl}{EMPIRE 2010}: CT lung registration
\item
  \href{https://masi.vuse.vanderbilt.edu/workshop2012/index.php/Main_Page}{Multi-Atlas
  Label Challenge 2012}: MRI brain registration and segmentation
\item
  \href{https://masi.vuse.vanderbilt.edu/workshop2013/index.php/MICCAI_2013_SATA_Challenge_and_Workshop:Current_events}{SATA
  Challenge 2013}: MRI cardiac and canine hind leg registration
\item
  \href{http://martinos.org/qtim/miccai2013/}{BRATS 2013}: Multi-modal
  MRI brain segmentation
\item
  \href{http://www.cardiacatlas.org/web/stacom2014/moco-introduction}{STACOM
  2014 MoCo Challenge}: MRI cardiac motion estimation
\end{itemize}

@@

\section{Major ANTs utilities}\label{major-ants-utilities}

\begin{frame}{Donoho?}

 \emph{``Papers are just advertisements for the science.''}

@@

\begin{frame}{Beyond original SyN}

\small

\includegraphics{./papers/figures/Frontiers_ITK.png}

\includegraphics{./papers/figures/Frontiers_BSplineSyN.png}

@@

\begin{frame}{Template building: creating the average Joe}

\includegraphics{./papers/figures/template0.jpg}

@@

\begin{frame}{``Attractiveness'' \(\rightarrow\) mental processing?}

\includegraphics{./papers/figures/template1.jpg}

@@

\begin{frame}{What about brains?}

\includegraphics{./papers/figures/template3.jpg}

@@

\begin{frame}{Templates facilitate computation}

\includegraphics{./papers/figures/template4.jpg}

@@

\begin{frame}{Nonparametric nonuniform intensity normalization (N3)}

\begin{itemize}
\item
  Developed at the Montreal Neurological Institute (John Sled, 1998)
\item
  Part of the standard preprocessing protocol in large scale projects
  such as ADNI
\item
  The traditional de facto standard in MRI bias correction

  \begin{itemize}
  \tightlist
  \item
    good performance
  \item
    \emph{public availability}
  \end{itemize}
\item
  Public availability --- set of perl scripts coordinating various C++
  programs
\item
  ``\emph{Let's incorporate N3 into ANTs!}''
\end{itemize}

@@

\begin{frame}[fragile]{N4 (``Nick's N3'')}

\begin{itemize}
\item
  comparative
  \href{http://www.ncbi.nlm.nih.gov/pubmed/20378467}{evaluation}
\item
  small spline distances (useful for higher magnet strengths)
\item
  multiresolution
\item
  weighted regional mask (used in
  \href{https://github.com/stnava/ANTs/blob/master/Scripts/antsAtroposN4.sh}{\texttt{antsAtroposN4.sh}})
\item
  fast execution times
\item
  \emph{publicly available}
\item
  tested nightly within the ITK dashboard system
\end{itemize}

@@

\begin{frame}{Multi-atlas segmentation}

\includegraphics{./tools/jointfusion/figures/jointLabelFusion.png}

@@

\begin{frame}{New work: joint intensity fusion}

\includegraphics{./tools/jointfusion/figures/jointIntensityFusion.png}

@@

\begin{frame}{Possible uses}

\begin{itemize}
\item
  ``Correct'' images

  \begin{itemize}
  \item
    motion correction
  \item
    ``remove'' lesions
  \end{itemize}
\item
  Project atlas set intensity signature
\item
  Use in ``corrective learning''
\end{itemize}

@@

\section{Putting it all together---the ANTs cortical thickness
pipeline}\label{putting-it-all-togetherthe-ants-cortical-thickness-pipeline}

\begin{frame}{Cortical thickness studies}

\begin{longtable}[c]{@{}ll@{}}
\toprule
Column1 & Column2\tabularnewline
\midrule
\endhead
Tetris-playing ability & chronic pancreatitis\tabularnewline
Huntington's disease & obsessive-compulsive disorder\tabularnewline
schizophrenia & ADHD\tabularnewline
bipolar disorder & obesity\tabularnewline
Alzheimer's disease & heritable depression\tabularnewline
frontotemporal dementia & elderly depression\tabularnewline
Parkinson's disease & age\tabularnewline
Williams syndrome & gender\tabularnewline
multiple sclerosis & handedness\tabularnewline
autism & intelligence\tabularnewline
migraines & athletic ability\tabularnewline
chronic smoking & meditative practices\tabularnewline
alcoholism & musical ability\tabularnewline
cocaine addiction & tendency toward criminality\tabularnewline
Tourette syndrome in children & childhood sexual abuse in female
adolescents\tabularnewline
scoliosis in female adolescents & traumatic brain injury\tabularnewline
early-onset blindness & untreated male-to-female
transsexuality\tabularnewline
\bottomrule
\end{longtable}

\begin{frame}{The ANTs structural brain mapping workflow}

\includegraphics{./evaluation/figures/pipeline.png}

\begin{frame}{Template building}

\emph{Tailor data to your specific cohort}

\includegraphics{./evaluation/figures/templates.png}

\begin{frame}{Template priors}

\includegraphics{./evaluation/figures/templatePriors.png}

\begin{frame}{Cortical thickness maps}

\includegraphics{./evaluation/figures/corticalThicknessEstimation.png}

In contrast to FreeSurfer which warps coupled surface meshes to segment
the gray matter, \emph{ANTs} diffeomorphically registers the white
matter to the combined gray/white matters while simultaneously
estimating thickness.

\begin{frame}

\emph{But without ground truth, how does one evaluate the pipeline?}

\begin{frame}{Predict age and gender}

\(AGE \sim VOLUME + GENDER + \sum_{i=1}^{62} T(DKT_i)\)

\begin{frame}{Prediction from cortical thickness data}

\includegraphics{./evaluation/figures/evaluation.png}

\section{Current work and Advanced Normalization Tools in R
(ANTsR)}\label{current-work-and-advanced-normalization-tools-in-r-antsr}

\begin{frame}{Multimodal Brain Tumor Segmentation (BRATS 2013)}

\includegraphics{./competitions/figures/brats2013results1.png}

\href{http://www.ncbi.nlm.nih.gov/pubmed/25433513}{Tustison, et al.,
Optimal symmetric multimodal templates and concatenated random forests
for supervised brain tumor segmentation (simplified) with \emph{ANTsR},
\emph{Neuroinformatics}.}

\begin{frame}{White matter hyperintensities in TBI}

\includegraphics{./wmhs/figures/wmhPipeline.png}

\begin{frame}{Social behavior and immunity dysfunction in mice}

\includegraphics{./antsr/figures/filiano_rsfmri.png}

\hypertarget{refs}{}

>>>>

\documentclass[ignorenonframetext,]{beamer}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\else % if luatex or xelatex
\ifxetex
\usepackage{mathspec}
\else
\usepackage{fontspec}
\fi
\defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
\usetheme{AnnArbor}
\usecolortheme{dolphin}
\usefonttheme{professionalfonts}
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\newif\ifbibliography
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight0.8\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}

@@

\section{Background}\label{background}

\begin{frame}{Founding developers}

BBA \& NJT

@@

\begin{frame}{Long-term collaborators}

\begin{figure}[htbp]
\centering
\includegraphics{./figures/antscollab.jpg}
\caption{}
\end{figure}

\(+\) \href{http://neuro.debian.net/pkgs/ants.html}{neurodebian},
\href{http://www.slicer.org/}{slicer},
\href{https://github.com/BRAINSia/BRAINSTools}{brainsfit},
\href{http://nipy.sourceforge.net/nipype/}{nipype},
\href{http://www.itk.org}{itk} and more \ldots{}

@@

\begin{frame}{}

\begin{figure}[htbp]
\centering
\includegraphics{./figures/lion.png}
\caption{}
\end{figure}

a pride: common way of doing things

\ldots{} in a competitive world \ldots{}

@@

\begin{frame}{Definitions}

\begin{itemize}
\item
  Registration \(=\) estimate an ``optimal'' geometric mapping between
  image pairs or image sets (e.g.~Affine)
\item
  Similarity \(=\) a function relating one image to another, given a
  transformation (e.g.~mutual information)
\item
  Diffeomorphisms \(=\) differentiable map with differentiable inverse
  (e.g. ``silly putty'', viscous fluid)
\item
  {Segmentation \(=\) labeling tissue or anatomy in images, usually
  automated (e.g.~K-means)}
\item
  Multivariate \(=\) using many voxels or measurements at once
  (e.g.~PCA, \(p >> n\) ridge regression)
\item
  {Multiple modality \(=\) using many modalities at once (e.g.~DTI and
  T1 and BOLD)}
\item
  MALF: multi-atlas label fusion - using anatomical dictionaries to
  label new data
\item
  {Solutions to challenging statistical image processing problems
  usually need elements from each of the above}
\end{itemize}

@@

\begin{frame}{Image mapping \& perception: 1878}

\begin{figure}[htbp]
\centering
\includegraphics{./figures/galton.png}
\caption{}
\end{figure}

\begin{itemize}
\item
  Francis Galton: Can we see criminality in the face?
\item
  (maybe he should have used \emph{ANTs}?)
\end{itemize}

@@

\begin{frame}{Image mapping \& biology: 1917}

\begin{figure}[htbp]
\centering
\includegraphics{./figures/dthompson.png}
\caption{}
\end{figure}

D'Arcy Thompson

@@

\begin{frame}{Initial scope}

\begin{figure}[htbp]
\centering
\includegraphics{./figures/ants_initial_scope.png}
\caption{}
\end{figure}

\ldots{} just do a better registration (tell story) \ldots{}

@@

\begin{frame}{\emph{ANTs} Lineage}

\begin{figure}[htbp]
\centering
\includegraphics{./figures/lineage.jpg}
\caption{}
\end{figure}

References: Horn and Schunck (1981), Gee, Reivich, and Bajcsy (1993),
Grenander (1993), Thompson et al. (2001), Miller, Trouve, and Younes
(2002), Shen and Davatzikos (2002), Arnold (2014), Thirion (1998),
Rueckert et al. (1999), Fischl (2012), Ashburner (2012)

@@

\begin{frame}{Diffeomorphisms}

plausible physical modeling of large, invertible deformations

``differentiable map with differentiable inverse''

@@

\begin{frame}{Fine-grained and flexible maps}

\begin{figure}[htbp]
\centering
\includegraphics{./figures/highresdiffeos.jpg}
\caption{}
\end{figure}

\ldots{} to correct a misconception about diffeomorphisms \ldots{}

@@

\begin{frame}{Diffeomorphisms: image parameterization in a metric space}

\begin{figure}[htbp]
\centering
\includegraphics{./figures/shape_appearance.png}
\caption{}
\end{figure}

@@

\begin{frame}[fragile]{General purpose library for multivariate image
registration, segmentation \& statistical analysis tools}

\begin{itemize}
\item
  170,000+ lines of C++, 6\(+\) years of work, 15+ collaborators.
\item
  Generic mathematical methods that are tunable for application specific
  domains: no-free lunch
\item
  Deep testing on multiple platforms \ldots{} osx, linux, windows.
\item
  Several ``wins'' in public knock-abouts (
  \href{http://www.ncbi.nlm.nih.gov/pubmed/19195496}{Klein 2009},
  \href{http://www.ncbi.nlm.nih.gov/pubmed/21632295}{Murphy 2011},
  \href{http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3837555/}{SATA 2012
  and 2013},
  \href{http://martinos.org/qtim/miccai2013/proc_brats_2013.pdf}{BRATS
  2013}, others )
\end{itemize}

\begin{verbatim}
    An algorithm must use prior knowledge about a problem
    to do well on that problem
\end{verbatim}

@@

\begin{frame}{\emph{ANTs}: Beyond Registration}

\begin{figure}[htbp]
\centering
\includegraphics{./figures/antscapabilities.jpg}
\caption{}
\end{figure}

\href{http://www.ncbi.nlm.nih.gov/pubmed/?term=atropos+tustison}{Atropos}
segmentation,
\href{http://www.ncbi.nlm.nih.gov/pubmed/?term=N4+tustison}{N4
inhomogeneity correction},
\href{http://www.ncbi.nlm.nih.gov/pubmed/?term=eigenanatomy+avants}{Eigenanatomy},
\href{http://www.ncbi.nlm.nih.gov/pubmed/?term=sparse+canonical+avants}{SCCAN},
\href{http://www.ncbi.nlm.nih.gov/pubmed/24852460}{Prior-constrained
PCA}, and
\href{http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4009425/}{atlas-based
label fusion} and
\href{http://www.ncbi.nlm.nih.gov/pubmed/21237273}{MALF} (powerful
expert systems for segmentation)

@@

\begin{frame}{On documentation}

\begin{figure}[htbp]
\centering
\includegraphics{figures/Chainsaw.jpg}
\caption{}
\end{figure}

documentation is important

@@

\begin{frame}{On documentation}

\begin{figure}[htbp]
\centering
\includegraphics{figures/blindmen.jpg}
\caption{}
\end{figure}

\ldots{} developers can be blind to doc deficiencies

while users are blind to what we provide!

\hypertarget{refs}{}
\hypertarget{ref-Arnold2014}{}
Arnold, Vladimir I. 2014. ``Topological Methods in Hydrodynamics.'' In
\emph{Vladimir I. Arnold-Collected Works}, 433--54. Springer.
\url{http://link.springer.com/chapter/10.1007/978-3-642-31031-7_42}.

\hypertarget{ref-Ashburner2012}{}
Ashburner, John. 2012. ``SPM: A History.'' \emph{Neuroimage} 62 (2).
Wellcome Trust Centre for Neuroimaging, 12 Queen Square, London WC1N
3BG, UK. john@fil.ion.ucl.ac.uk: 791--800.
doi:\href{https://doi.org/10.1016/j.neuroimage.2011.10.025}{10.1016/j.neuroimage.2011.10.025}.

\hypertarget{ref-Fischl2012}{}
Fischl, Bruce. 2012. ``FreeSurfer.'' \emph{Neuroimage} 62 (2). Athinoula
A Martinos Center, Dept. of Radiology, MGH, Harvard Medical School, MA
fischl@nmr.mgh.harvard.edu, USA. fischl@nmr.mgh.harvard.edu: 774--81.
doi:\href{https://doi.org/10.1016/j.neuroimage.2012.01.021}{10.1016/j.neuroimage.2012.01.021}.

\hypertarget{ref-Gee1993}{}
Gee, J. C., M. Reivich, and R. Bajcsy. 1993. ``Elastically Deforming 3D
Atlas to Match Anatomical Brain Images.'' \emph{J Comput Assist Tomogr}
17 (2). Department of Computer; Information Science, University of
Pennsylvania, Philadelphia 19104.: 225--36.

\hypertarget{ref-Grenander1993}{}
Grenander, Ulf. 1993. \emph{General Pattern Theory-a Mathematical Study
of Regular Structures}. Clarendon Press.

\hypertarget{ref-Horn1981}{}
Horn, Berthold K, and Brian G Schunck. 1981. ``Determining Optical
Flow.'' In \emph{1981 Technical Symposium East}, 319--31. International
Society for Optics; Photonics.
\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=1231385}.

\hypertarget{ref-Miller2002}{}
Miller, Michael I., Alain Trouve, and Laurent Younes. 2002. ``On the
Metrics and Euler-Lagrange Equations of Computational Anatomy.''
\emph{Annu Rev Biomed Eng} 4. Center for Imaging Science, The Johns
Hopkins University, Baltimore, Maryland 21218, USA. mim@cis.jhu.edu:
375--405.
doi:\href{https://doi.org/10.1146/annurev.bioeng.4.092101.125733}{10.1146/annurev.bioeng.4.092101.125733}.

\hypertarget{ref-Rueckert1999}{}
Rueckert, D., L. I. Sonoda, C. Hayes, D. L. Hill, M. O. Leach, and D. J.
Hawkes. 1999. ``Nonrigid Registration Using Free-Form Deformations:
Application to Breast MR Images.'' \emph{IEEE Trans Med Imaging} 18 (8).
Division of Radiological Sciences; Medical Engineering, Guy's, King's,;
St. Thomas' School of Medicine, King's College London, Guy's Hospital,
London, UK. D.Rueckert@umds.ac.uk: 712--21.
doi:\href{https://doi.org/10.1109/42.796284}{10.1109/42.796284}.

\hypertarget{ref-Shen2002}{}
Shen, Dinggang, and Christos Davatzikos. 2002. ``HAMMER: Hierarchical
Attribute Matching Mechanism for Elastic Registration.'' \emph{IEEE
Trans Med Imaging} 21 (11). Center for Biomedical Image Computing,
Department of Radiology, The Johns Hopkins University School of
Medicine, 601 N. Caroline Street, Baltimore, MD 21287, USA.
dgshen@rad.upenn.edu: 1421--39.
doi:\href{https://doi.org/10.1109/TMI.2002.803111}{10.1109/TMI.2002.803111}.

\hypertarget{ref-Thirion1998}{}
Thirion, J. P. 1998. ``Image Matching as a Diffusion Process: An Analogy
with Maxwell's Demons.'' \emph{Med Image Anal} 2 (3). INRIA, Equipe
Epidaure, Sophia-Antipolis, France. jean-philippe.thirion@inria.fr:
243--60.

\hypertarget{ref-Thompson2001}{}
Thompson, Paul M., Michael S. Mega, Christine Vidal, Judith L. Rapoport,
and Arthur W. Toga. 2001. ``Detecting Disease-Specific Patterns of Brain
Structure Using Cortical Pattern Matching and a Population-Based
Probabilistic Brain Atlas.'' \emph{Inf Process Med Imaging} 2082.
Laboratory of Neuro Imaging, Division of Brain Mapping,; Alzheimer's
Disease Center, Dept. of Neurology, UCLA School of Medicine, Los
Angeles, CA, USA thompson@loni.ucla.edu.: 488--501.
doi:\href{https://doi.org/10.1007/3-540-45729-1_52}{10.1007/3-540-45729-1\_52}.
